---
title: "BANA482 Case 3: Time Series Forecasting"
author: "Mohammed Baobaid - 202031137"
date: "2026-02-18"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE,
  message = FALSE
)
```

# 0. Setup & Reproducibility

## 0.1 Install Required Packages (Run Once)

```{r install-packages, eval=FALSE}
required_packages <- c(
  "tidyverse",
  "lubridate",
  "forecast",
  "tseries"
)

installed <- rownames(installed.packages())

for (pkg in required_packages) {
  if (!(pkg %in% installed)) {
    install.packages(pkg)
  }
}
```

## 0.2 Load Libraries & Set Seed

```{r load-libraries}
SEED <- 202031137
set.seed(SEED)

library(tidyverse)
library(lubridate)
library(forecast)
library(tseries)
```

# 1. Data Loading and Initial Inspection

## 1.1 Load Dataset

```{r load-data}

file_path <- "../data/train.csv"

if (!file.exists(file_path)) {
  stop("File not found. Please check dataset path.")
}

train <- read_csv(file_path, show_col_types = FALSE)

head(train)
```

## 1.2 Structure & Quality Check

```{r inspect-data}

glimpse(train)
summary(train)
colSums(is.na(train))
```

# 2. Data Preparation

## 2.1 Convert Date Column

```{r convert-date}

train <- train %>%
  mutate(date = as.Date(date))

if (any(is.na(train$date))) {
  stop("Date conversion produced missing values. Please check the date format.")
}
```

## 2.2 Filter to Single Store & Item

```{r filter-series}

train_filtered <- train %>%
  filter(store == 1, item == 1)

nrow(train_filtered)
```

## 2.3 Aggregate Daily to Monthly Sales

```{r aggregate-monthly}

monthly_sales <- train_filtered %>%
  mutate(month = floor_date(date, "month")) %>%
  group_by(month) %>%
  summarise(monthly_sales = sum(sales), .groups = "drop")

head(monthly_sales)
```

## 2.4 Create Time Series Object

```{r create-ts}

sales_ts <- ts(
  monthly_sales$monthly_sales,
  start = c(year(min(monthly_sales$month)),
            month(min(monthly_sales$month))),
  frequency = 12
)

sales_ts
```

## 2.5 Initial Time Series Plot

```{r plot-ts}

autoplot(sales_ts) +
  ggtitle("Monthly Sales Time Series (Store 1, Item 1)") +
  xlab("Year") +
  ylab("Units Sold")
```

# 3. Exploratory Time Series Analysis

## 3.1 Seasonal Pattern

```{r seasonal-plot}

ggseasonplot(sales_ts, year.labels = TRUE) +
  ggtitle("Seasonal Pattern of Monthly Sales")
```

## 3.2 Decomposition

```{r decomposition}

decomp_add <- decompose(sales_ts, type = "additive")
decomp_mult <- decompose(sales_ts, type = "multiplicative")

autoplot(decomp_add) + ggtitle("Additive Decomposition")
autoplot(decomp_mult) + ggtitle("Multiplicative Decomposition")
```

## 3.3 Compare Decomposition Fit

```{r decomposition-sd}

sd(decomp_add$random, na.rm = TRUE)
sd(decomp_mult$random, na.rm = TRUE)
```

# 4. Stationarity Testing

## 4.1 ADF Test

```{r adf-test}

adf.test(sales_ts)
```

## 4.2 KPSS Test

```{r kpss-test}

kpss.test(sales_ts)
```

## 4.3 Differencing

```{r differencing}

diff_1 <- diff(sales_ts, differences = 1)
autoplot(diff_1) + ggtitle("First-Order Differencing")

diff_seasonal <- diff(sales_ts, lag = 12)
autoplot(diff_seasonal) + ggtitle("Seasonal Differencing")
```

# 5. Model Development

## 5.1 ARIMA Model

```{r fit-arima}

arima_model <- auto.arima(
  sales_ts,
  seasonal = TRUE,
  stepwise = FALSE,
  approximation = FALSE
)

summary(arima_model)
```

## 5.2 Residual Diagnostics

```{r arima-diagnostics}

checkresiduals(arima_model)
```

## 5.3 ETS Model

```{r fit-ets}

ets_model <- ets(sales_ts)
summary(ets_model)
```

## 5.4 Model Comparison

```{r compare-models}

accuracy(arima_model)
accuracy(ets_model)
```

# 6. Forecasting

## 6.1 ARIMA Forecast

```{r arima-forecast}

arima_forecast <- forecast(
  arima_model,
  h = 6,
  level = c(80, 95)
)

autoplot(arima_forecast) +
  ggtitle("6-Month Forecast (ARIMA)")
```

## 6.2 ETS Forecast

```{r ets-forecast}

ets_forecast <- forecast(
  ets_model,
  h = 6,
  level = c(80, 95)
)

autoplot(ets_forecast) +
  ggtitle("6-Month Forecast (ETS)")
```

## 6.3 Forecast Table

```{r forecast-table}

forecast_table <- data.frame(
  Month = time(arima_forecast$mean),
  Forecast = as.numeric(arima_forecast$mean),
  Lower_80 = arima_forecast$lower[,1],
  Upper_80 = arima_forecast$upper[,1],
  Lower_95 = arima_forecast$lower[,2],
  Upper_95 = arima_forecast$upper[,2]
)

forecast_table
```

# 7. Model Evaluation

## 7.1 Train-Test Split

```{r split-data}

train_ts <- head(sales_ts, length(sales_ts) - 6)
test_ts  <- tail(sales_ts, 6)
```

## 7.2 Refit Models

```{r refit-models}

arima_train <- auto.arima(train_ts)
ets_train   <- ets(train_ts)
```

## 7.3 Forecast Test Period

```{r forecast-test}

arima_test_forecast <- forecast(arima_train, h = 6)
ets_test_forecast   <- forecast(ets_train, h = 6)
```

## 7.4 Accuracy Metrics

```{r evaluate-models}

accuracy(arima_test_forecast, test_ts)
accuracy(ets_test_forecast, test_ts)

actual <- as.numeric(test_ts)

predicted_arima <- as.numeric(arima_test_forecast$mean)
mape_arima <- mean(abs((actual - predicted_arima) / actual)) * 100
rmse_arima <- sqrt(mean((actual - predicted_arima)^2))

predicted_ets <- as.numeric(ets_test_forecast$mean)
mape_ets <- mean(abs((actual - predicted_ets) / actual)) * 100
rmse_ets <- sqrt(mean((actual - predicted_ets)^2))

mape_arima
rmse_arima
mape_ets
rmse_ets
```
